---
title: "STAT 350 (Survival Analysis): Final group assignment"
author: "Shamil Khusnullin"
date: "2026-04-26"
date-format: iso
format: pdf
toc: true
editor: source
---

```{r}
#| label: setup
#| include: false
library(mosaic)   
library(tidyverse)
library(survival)
library(rsurv)
rnd_val <- 2
```

## Instructions

This final group assignment has two parts:

1. find a journal article (ideally in your topic area, but that's not necessary) that fits and reports a Cox proportional hazards model.
Add the pdf of the article to the `assignment_2` folder, then review how the authors report and assessed the Cox model.
2. Carry out a power calculation for an investigator planning a study with a survival time outcome.
Utilize a simulation study to estimate the power needed given the scenario provided and the sample size available.

This assignment is due via GitHub submission by the end of the day on Wednesday, April 29th.

> Please indicate any resources such as generative AI beyond the instructor, textbook, notes, or materials from the class that were used in your solution (or indicate "We used no additional materials or resources"):

I used google's automatic AI-generated response (the first tab) when looking up syntaxes of code and no other additional resources. 

\newpage

## Cox model in practice

Find a journal article that fits and reports a Cox proportional hazards model.
Share the abstract and title page with Nick (using the issue that he has created for you).
Once Nick has signed off, you should  then review how the authors report and assessed the Cox model in their paper.
Include a in-text citation (no BibTeX needed!) for the article, provide a brief summary of the article, describe how the authors assessed the Cox model, and share any comments you have on their approach.

SOLUTION:

- TITLE: An Application of the Cox Proportional Hazards Model to Bank Failure
- AUTHORS: William R. Lane, Stephen W. Looney, and James W. Wansley
- JOURNAL: Journal of Banking and Finance
- YEAR: 1986
- VOLUME: 10, Issue 4, pages 511–531
- URL/DOI: 10.1016/S0378-4266(86)80003-6

XX



\newpage

## Power calculation introduction

One of your classmates has gotten a degree in public health and has a new job working in a large research lab.
They have been tasked with coming up with an estimate of power for a survival study they are planning.

Scenario A: Here are the details of the study:

1. They can only afford a total of n = 300 subjects (150 in each of two treatment groups).
2. They anticipate that the underlying survival times are Weibull with scale parameter = 5 and shape parameter = 3 (using the `rsurv::rphreg()` parametrization).
3. The true log hazard rate (comparing treatment to control) is assumed to be -0.4 (i.e., a hazard ratio of exp(-0.4) = 0.67).
4. They will follow subjects for a maximum of five (5) weeks (time unit in the study is weeks).
5. The subjects are enrolled at the start of the study (i.e., there is no staggered entry) and are followed until the end of the study or until they have an event, whichever comes first.
6. The survival times are independent across subjects (so no frailties are needed).
7. The investigators will analyze the data using a logrank test to compare the two treatment groups.

Carry out an empirical power calculation for this study with `num_sim = 2000` simulations, assuming that the investigators will analyze the data using a logrank test.

(Hint 1: to check that your results make sense, you may want to first run a simulation with the treatment effect set to 0, which should give you an estimate of power close to 0.05!)

(Hint 2: the `binom.test()` function can be used to calculate a confidence interval around your estimate of power, which is a convenient way to quantify the uncertainty in your estimate.)

### Scenario A

## Code and simulations

```{r}
set.seed(1821)
num_sim <- 2000 

gen_data <- function(
  beta_1 = -0.4, # log hazard ratio for treatment effect
  itt_prob = 0.0, # P(intervention subject doesn't take meds)
  shape_val = 3, # underlying shape value
  scale_val = 5, # underlying scale value
  n = 300, # total sample size (half randomized to each group)
  tau = 5 # maximum follow up time
    ) {
  simulated_data <- tibble(
    randomized_treat = rep(c(0, 1), times = n/2)
  )|>
    mutate(
      actual_treat = if_else(
        randomized_treat == 1,
        rbinom(n(), size = 1, prob = 1 - itt_prob),
        0L
      ),
      true_t = rphreg(
        runif(n),
        ~ actual_treat,
        beta = beta_1,
        dist = "weibull",
        shape = shape_val,
        scale = scale_val
      ),
      time = pmin(true_t, tau),
      status = as.numeric(true_t <= tau)
    ) |>
    select(-true_t)

  return(simulated_data)
}

glimpse(gen_data())

  

```


```{r}
#| eval: true
run_sim0 <- function() {
  data <- gen_data(beta_1 = 0)
  fit <- survdiff(Surv(time, status) ~ randomized_treat, data = data)
  return(fit$pvalue)
}

tictoc::tic()
results0 <- 1:num_sim |>
  purrr::map_dbl(~ run_sim0())
binom.test(~ results0 <= 0.05) # matches what we would expect!
tictoc::toc()
```

As a check, I first ran the simulation with no treatment effect. The estimated rejection rate was close to 0.05, which is what I would expect under the null hypothesis. 

```{r}
set.seed(1821)
data_A_example <- gen_data(beta_1 = -0.4, itt_prob = 0)
mean(data_A_example$status == 0)

S_control <- exp(-(5 / 5)^3)
S_treat <- exp(-exp(-0.4) * (5 / 5)^3)
mean_censored_A <- mean(c(S_control, S_treat))
mean_censored_A

```


\noindent
a) Approximately what proportion of the subjects are censored in this study? (Hint: you can get a quick estimate by simulating a single dataset and looking at the proportion of censored subjects, but you can also calculate this using the underlying distribution and parameters.)

SOLUTION: 44% are censored. In fact, as there is no censorship that happens during the follow up time and the censorhip only happens for patients that survived for longer than 5 weeks - the censorhip rate is just mean of survived patients among 2 groups. 

\noindent

```{r}
set.seed(1821)
fit_fun <- function(beta_1 = -0.4, itt_prob = 0) {
  data <- gen_data(beta_1 = beta_1, itt_prob = itt_prob)

  fit <- survdiff(Surv(time, status) ~ randomized_treat, data = data)

  p_val <- 1 - pchisq(fit$chisq, df = 1)

  tibble(
    beta_1 = beta_1,
    itt_prob = itt_prob,
    Wald.p = p_val
  )
}

results_A <- 1:num_sim |>
  purrr::map_dfr(~ fit_fun(beta_1 = -0.4, itt_prob = 0)) |>
  mutate(reject = Wald.p <= 0.05)

binom.test(sum(results_A$reject), num_sim)

```

b) Report your estimate of power for this study, along with a 95% confidence interval around your estimate.
Interpret your findings.

The estimated power was approximately 0.732, with a 95% confidence interval of (0.71 0.75). This means that, if the true log hazard ratio is -0.4, the study would reject the null hypothesis about 73.2% of the time at the 0.05 significance level.



\newpage

### Scenario B

Scenario B: Your classmate comes back and says that they are concerned about non-compliance in the treatment group.

Based on pilot data, they anticipate that 10% of subjects randomized to the treatment group will not take their medication (i.e., they will be "non-compliant" and have the same hazard as the control group).
These subjects will be analyzed in their randomized treatment group, but the simulation should assume that they are just like the controls.

Repeat the power calculation for the study under Scenario B (again calculating and reporting a 95% confidence interval around your estimate of power), and interpret your findings.


```{r}
results_B <- 1:num_sim |>
  purrr::map_dfr(~ fit_fun(beta_1 = -0.4, itt_prob = 0.1)) |>
  mutate(reject = Wald.p <= 0.05)

binom.test(sum(results_B$reject), num_sim)


```

SOLUTION FOR SCENARIO B:

In Scenario B, 10% of subjects randomized to treatment are assumed not to take the medication, so their event times are generated as if they were controls. However, they are still analyzed in the treatment group. The estimated power was approximately 0.64, with a 95% confidence interval of (0.62, 0.66). Compared with Scenario A, the power is lower because non-compliance weakens the observed difference between the randomized treatment and control groups. 


### Conclusion

Write a short (no more than four paragraph) conclusion summarizing your findings from the power calculations in Scenario A and B, and any comments you have on the impact of non-compliance on power in this setting.

SOLUTION:

In Scenario A, where all subjects randomized to treatment are assumed to actually receive the treatment, the simulation suggests that the planned study has moderate power to detect the assumed treatment effect. With a true log hazard ratio of -0.4, corresponding to a hazard ratio of about 0.67, the estimated power was approximately 0.732, with a 95% ci of about (0.71, 0.75). This means that under these assumption and confidence level of 0.05 we're going to reject the null hypothesis (where null hypothesis is 0 treatment effect) only in 73.2%

In Scenario B, where 10% of subjects randomized to treatment are assumed not to take the medication, the estimated power decreased to approximately 0.64, with a 95% ci of (0.62, 0.66). This drop makes sense because the treatment group now contains some subjects whose event times behave like the control group, even though they are still analyzed as part of the treatment group. As a result, the observed effect of treatment in the treatment group is lower than it's actually is. 

Overall, these simulations show that non-compliance can meaningfully reduce statistical power even when the sample size and follow-up time stay the same. In this setting, the planned study already has only moderate power under perfect compliance, and the introduction of 10% non-compliance makes the study less likely to detect the true treatment effect.

